---
title: "Bondone_script"
output: html_document
date: "2026-02-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
install.packages('tidyverse')
library(tidyverse)
install.packages('phyloseq') #not working

library(readr)

install.packages('iCAMP')
library(iCAMP)

library(vegan)
library(ade4)


```

#install phyloseq
```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("phyloseq")
library(phyloseq)

```

```{r}
library(Biostrings)
```

####################
# *Import the files*

# import tree
```{r}
Bact_tree<-read_tree('16S/16s_bacterialtree.nwk')
```
# import fasta (this actually is not needed by phyloseq)
```{r}
Bact_fasta<-readDNAStringSet('16S/16s_bacteria_only.fasta')
```

# import OTU
```{r}
Bact_otu<-read.csv('16S/otu_table_bacteria.csv')

rownames(Bact_otu)<-Bact_otu$SeqName # assign rownames
Bact_otu$X<-NULL
Bact_otu$SeqName<-NULL

# remove the 'X' in front of sample names
colnames(Bact_otu)<- gsub( 'X', '', colnames(Bact_otu) )
```

#import sample tab
```{r}
sample_tab<-read.csv('16S/sample_table.csv')

rownames(sample_tab)<-sample_tab$uniqueID # assign rownames
sample_tab$X<-NULL
```

#import tax table
```{r}
Bact_tax<-read.csv('16S/tax_table_bacteria.csv')

rownames(Bact_tax)<-Bact_tax$SeqName # assign row names 
Bact_tax<-Bact_tax[,-c(1,2)]

```

# *import all into phyloseq*

```{r}
class(Bact_tax)
Bact_tax<-as.matrix(Bact_tax) # convert to matrix
```

# create the needed dataframes for phyloseq
```{r}
b.OTU<-otu_table(Bact_otu, taxa_are_rows = T)
b.TAX<-tax_table(Bact_tax)
b.sample<-sample_data(sample_tab)

```

# create the phyloseq file *b_ps* - all OTUs for now (will subset later)
```{r}
b_ps<-phyloseq(b.OTU, b.TAX, b.sample, Bact_tree) # this includes all otus

```

# *examine the OTU tab - check abundance, occurrences, etc*
# plot the abundance and occurrence (richness) profile of taxa per sample
*discuss if rarefying to 12k reads is needed* - 
```{r}
sort.abb<-sort(colSums(Bact_otu), decreasing = T) # sort tot abundance per sample
plot(sort.abb, type='l')

sort.occ<-sort(colSums(vegan::decostand(Bact_otu, 'pa')), decreasing = T) # sort tot richness per sample
plot(sort.occ, type='l')


# many taxa found in only 1 sample
sort_taxa_occur<-sort(rowSums(vegan::decostand(Bact_otu, 'pa'))) # sort tot occurrence per taxa (in how many samples does the taxa occur)
plot(sort_taxa_occur, type='l')

sum(sort_taxa_occur>1) # many with low occurrence


sort_taxa_abb<-sort(rowSums(Bact_otu)) # sort tot abundance per taxa
plot(sort_taxa_abb, type='l')

sum(sort_taxa_abb>5)


```


# Working on the *niche differences between taxa*
```{r}
#traspose the Otu tab
Bact_otu_t<-t(Bact_otu)
Bact_otu_t<-as.data.frame(Bact_otu_t)
```

# niche difference between OTUs (all OTUs now) bases on elevation (needed for Mantel test, etc)
```{r}
niche.dif=iCAMP::dniche(env = sample_tab[,c(2,4,5,6)], 
                        comm =Bact_otu_t,
                        method = "niche.value",
                        nworker = 4,out.dist=T, bigmemo=F,
                        nd.wd="~/Documents/FEM_work/PNRR/Bondone")
```

# check the mean and variance of each taxa over the elevation gradient- elevetion niche (using all taxa)
```{r}
elev_niche=
ade4::sco.distri(sample_tab[,2],
           Bact_otu_t ,
            labels = NULL)
```

```{r}
class(Bact_tax)
Bact_tax<-as.data.frame(Bact_tax)

```

# add phylum and Otu info on the elev_niche df
```{r}
elev_niche$Phylum= Bact_tax$Phylum[match(row.names(elev_niche), row.names(Bact_tax))]
elev_niche$OTU= rownames(elev_niche)


```


# plot elevation niche using all taxa
```{r fig.width=6, fig.height=9}
 ggplot(data=elev_niche ,
        aes(x=mean, y=reorder(OTU, mean), xmin=(mean-sqrt(var)), xmax=(mean+sqrt(var))))+xlab('Elevation')+ylab('Bacterial taxa')+
    geom_point(data=elev_niche %>% select(mean, var, OTU), aes(x=mean, y=reorder(OTU, mean), xmin=(mean-sqrt(var)), xmax=(mean+sqrt(var))),   size=1, shape=21, col='grey60')+geom_errorbarh(size=0.3, col='grey60')+theme(axis.text.x = element_text(angle=45, vjust=0, hjust=0))+xlim(700, 1650)+

theme_classic()+
  
    geom_point(aes(col=Phylum ))+
 geom_errorbarh(aes(col=Phylum), size=0.3)+theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())+facet_wrap(~Phylum)+
  theme(axis.text.x = element_text(angle=45, vjust=0, hjust=0), axis.title = element_text(size=15))+theme(legend.position = 'none')
```

#subsetting
# *subset otu table excluding taxa present in only 3 sample*

```{r}
Bact_otu_subs<-
Bact_otu %>% 
  filter(rownames(Bact_otu) %in% names(sort_taxa_occur[sort_taxa_occur>3]) ) # filter using names from the df 'sort_taxa_occur'

# these the names of taxa occurring in >3 sample
names(sort_taxa_occur[sort_taxa_occur>3])

# how many reads per OTU in the subsetted group?
sort_taxa_abb_subs<-sort(rowSums(Bact_otu_subs)) # within the taxa occurring in >3 samples, the min read (abb) is 5

```
# Traspose subsetted OTU tab
```{r}
Bact_otu_subs_t<-as.data.frame(t(Bact_otu_subs))
```




# calculate niche elevation, using the subset of taxa
```{r}
elev_niche_subs=
sco.distri(sample_tab[,2],
           Bact_otu_subs_t ,
            labels = NULL)
```
# add tax information
```{r}
elev_niche_subs$Phylum= Bact_tax$Phylum[match(row.names(elev_niche_subs), row.names(Bact_tax))]
elev_niche_subs$OTU= rownames(elev_niche_subs)
```

# plot elevation niche with subset of taxa (c 11k otus)
```{r fig.width=6.5, fig.height=9}
 ggplot(data=elev_niche_subs ,
        aes(x=mean, y=reorder(OTU, mean), xmin=(mean-sqrt(var)), xmax=(mean+sqrt(var))))+xlab('Elevation')+ylab('Bacterial taxa')+
    geom_point(data=elev_niche_subs %>% select(mean, var, OTU), aes(x=mean, y=reorder(OTU, mean), xmin=(mean-sqrt(var)), xmax=(mean+sqrt(var))),   size=1, shape=21, col='grey60')+geom_errorbarh(size=0.3, col='grey60')+theme(axis.text.x = element_text(angle=45, vjust=0, hjust=0))+xlim(700, 1650)+

theme_classic()+
  
    geom_point(aes(col=Phylum ))+
 geom_errorbarh(aes(col=Phylum), size=0.3)+theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())+facet_wrap(~Phylum)+
  theme(axis.text.x = element_text(angle=45, vjust=0, hjust=0), axis.title = element_text(size=15), strip.text = element_text(size=7))+theme(legend.position = 'none')
```


## *Update phyloseq objet, using subset of taxa*

```{r}
b_ps_subs<-prune_taxa(taxa_names(b_ps) %in% names(Bact_otu_subs_t), b_ps )
```

# calculate the *niche difference* using subsetted phyloseq (based on 6631 otu)
```{r}
niche.dif_subs=iCAMP::dniche(env = sample_tab[,c(2,4,5,6)], 
                        comm =Bact_otu_subs_t,
                        method = "niche.value",
                        nworker = 4,out.dist=T, bigmemo=F,
                        nd.wd="~/Documents/FEM_work/PNRR/Bondone")
```

# extract the phylogen distance *pd.subs*
```{r}
pd.subs=cophenetic(phy_tree(b_ps_subs))
```

# Plot the mantel correlogram for altitude
*this takes about 10m*
```{r}
mantest.elev=
mantel.correlog(as.dist(pd.subs),as.dist(niche.dif_subs$nd$altitude), nperm = 150 )
```

```{r}
plot(mantest.elev, alpha=.05) 
title('Correlogram Elevation')
```

# transform the elevation and phylo distance into 3cols matrices
```{r}
elev_dist<-as.dist(niche.dif_subs$nd$altitude)
pd_dist<-as.dist(pd.subs)

elev_dist_3col<-
dist.3col(elev_dist)

elev_d

pd_dist_3col<-
  dist.3col(pd_dist)



```

# add the otu-pairs to each df
```{r}
pd_dist_3col$pairs<-paste(pd_dist_3col$name1, pd_dist_3col$name2, sep='_')

elev_dist_3col$pairs<-paste(elev_dist_3col$name1, elev_dist_3col$name2, sep='_')

```

# add the phylo distance into the elevation distance df (using the same otu pairs to guide the match)
```{r}
elev_dist_3col$pd_dist<-pd_dist_3col$dis[match(elev_dist_3col$pairs, pd_dist_3col$pairs )]

```

# explore relation between raw distances (elev vs phylo)
```{r}
elev_dist_3col %>% 
  ggplot()+aes(pd_dist, dis)+geom_smooth()
```


